---
title: "Exploratory Data Analysis (EDA) - IEEE-CIS Fraud Detection"
author: "Alejandro Matías Astorga"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# 1.Introduction

In this notebook, we perform an exploratory data analysis (EDA) on the IEEE-CIS Fraud Detection dataset. The goal is to understand the characteristics of fraudulent and non-fraudulent transactions, identify patterns, and prepare for machine learning modeling.

# 2. Libraries

```{r libraries}
library(tidyverse)
library(data.table)
library(ggplot2)
library(gridExtra) # For multiple plots
```

# 3. Data Loading


```{r data-loading}
library(tidyverse)
library(data.table)
library(ggplot2)
library(gridExtra) # For multiple plots
```


# 4. Data Overview

## 4.1 Missing Values

```{r missing-values}
# Percentage of missing values per column
missing_data <- train_data %>%
  summarise_all(~ mean(is.na(.))) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "missing_perc") %>%
  arrange(desc(missing_perc))

head(missing_data, 20)
```


## 4.2 Class Balance

```{r class-balance}
# Distribution of target variable
train_data %>%
  count(isFraud) %>%
  mutate(percentage = n / sum(n))
```

# 5. Exploratory Visualizations

## 5.1 Transaction Amount Distribution
```{r transaction-amound}
ggplot(train_data, aes(x = TransactionAmt, fill = factor(isFraud))) +
  geom_histogram(bins = 100, position = "identity", alpha = 0.6) +
  scale_x_log10() + 
  labs(title = "Transaction Amount Distribution (Log Scale)", fill = "Is Fraud")
```

## 5.2 Device Info

```{r transaction-amound}
train_data %>%
  count(DeviceInfo) %>%
  top_n(10, wt = n) %>%
  ggplot(aes(x = reorder(DeviceInfo, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Device Info", x = "Device Info", y = "Count")
```

# 6. Conclusions and Insights
- Class Imbalance: Describe the fraud ratio.
- Key Patterns: Summarize key insights from visualizations.
- Next Steps: Prepare for feature engineering and modeling.